Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2025-05-25 manuscript/manuscript.tex
Working on discussion

The models in this study can assist physicians in assigning CODs in a variety of ways due to low costs and speed of COD assignment. Similar to differential diagnoses, GPT and InSilicoVA models offer alternative COD assignments for physicians to consider \citep{barnettComparativeAccuracyDiagnosis2019}, which can potentially help lower the number of records with ill-defined causes or reduce disagreement between physicians. At the time of this study, running GPT-3.5 cost $\sim$\$1.6 USD (\$0.5 per one million tokens), GPT-4 cost $\sim$\$115 USD (\$30 per one million tokens), and InSilicoVA was cost free on 6939 records \citep{openaiPricing2024}. These costs were lower than physicians (e.g. less than \$3 USD per house in India \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005}), while it is possible to code over 10,000 records in under a day. When physicians are unavailable, GPT and InSilicoVA models can be a cost-efficient alternative to code large amounts of records for population estimates of CODs. However, it is recommended to apply these models only for certain CODs where models perform well, such as in Table \ref{tab-leadcod}. In addition, these models can also help divert physician resources to cases that are more difficult to code or require more attention. For example, physicians can validate cases where models performed well (e.g. maternal conditions at 0.79-0.99 PCCC), while spending more time on cases where models performed poorly (e.g. acute respiratory infections at 0.25-0.61 PCCC). Lastly, as models can assign CODs on-demand, there is potential for models to provide immediate COD estimates during the data collection process to monitor data quality (e.g. comparing estimated to expected COD distributions for known areas as quality checks).

@@

All models did not require training data to assign CODs, which allowed them to be used without domain expertise and supplying training datasets. The main advantage to GPT-3.5 and GPT-4 was the use of natural language text as input and output. Compared to InterVA-5 and InSilicoVA, GPT models were able to assign COD codes in ICD-10, as physicians do, and potentially assign CODs in more broad categories depending on the prompts. In comparison, InterVA-5 and InSilicoVA relied on structured input and output data from WHO VA 2016 questionnaires, and assigned CODs in WHO VA 2016 codes only. This required that these codes and forms be maintained with conversions between different form (e.g. WHO VA 2012 to WHO VA 2016) and code standards (e.g. WHO VA 2016 to ICD-10), which reduces interoperability and comparability when other models and VA systems do not use the same or interchangeable standards. Thus, GPT models did not require strict formats for training and testing data, which can capture latent and more ambiguous patterns (e.g. health-seeking behaviours and social issues) outside the scope of WHO VA codes and forms \citep{jebleeAutomaticallyDeterminingCause2019,kingQualityDiagnosticValue2016}. For example, GPT-3.5 and GPT-4 had higher performance (+0.13-0.68 PCCC) than InterVA-5 and InSilicoVA for more ambiguous CODs (e.g. other/unspecified infections, ill-defined, other cardiovascular diseases). GPT models also performed better (+0.57 PCCC) on CODs with a small number of cases, such as nutritional deficiencies (n=11) and diabetes mellitus (n=27), which may not have enough cases for questionnaires to capture statistical patterns, but may possibly have richer contextual information from articles, web sources, or books that GPT models can leverage. GPT models may also guide surveyors during VA interviews to ask better questions from households that improve narrative quality.

@@

This section discusses research opportunities to improve GPT models for assigning CODs. Section \ref{discuss-opp-peng} discusses the potential to improve GPT models with prompt engineering and exploration of misclassified records. Section \ref{discuss-opp-vaint} identifies an opportunity to integrate GPT, InterVA-5, and InSilicoVA models into VA systems for improving physician COD assignment.

@@

\subsubsection{Computer Assisted Verbal Autopsy}\label{discuss-opp-vaint}

>>>>

The models in this study can assist physicians in assigning CODs in a variety of ways due to low costs and speed of COD assignment. Similar to differential diagnoses, GPT and InSilicoVA models offer alternative COD assignments for physicians to consider \citep{barnettComparativeAccuracyDiagnosis2019}, which can potentially help lower the number of records with ill-defined causes or reduce disagreement between physicians. At the time of this study, running GPT-3.5 cost $\sim$\$1.6 USD (\$0.5 per one million tokens), GPT-4 cost $\sim$\$115 USD (\$30 per one million tokens), and InSilicoVA was cost free on 6939 records \citep{openaiPricing2024}. These costs were lower than physicians (e.g. less than \$3 USD per house in India \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005}), while it is possible to code over 10,000 records in under a day. When physicians are unavailable, GPT and InSilicoVA models can be a cost-efficient alternative to code large amounts of records for population estimates of CODs. However, it is recommended to apply these models only for certain CODs where models perform well, such as in Table \ref{tab-leadcod}. In addition, these models can also help divert physician resources to cases that are more difficult to code or require more attention. For example, physicians can validate cases where models performed well (e.g. maternal conditions at 0.79-0.99 PCCC), while spending more time on cases where models performed poorly (e.g. acute respiratory infections at 0.25-0.61 PCCC).

@@

Training data was not required to assign CODs for all models, which allowed application without domain expertise or supplying training datasets. The main advantage to GPT-3.5/4 was the use of natural language text as input and output. Compared to InterVA-5 and InSilicoVA, GPT models were able to assign COD codes in ICD-10 standard, as physicians do, and potentially assign CODs in more broad categories depending on the prompts. In comparison, InterVA-5 and InSilicoVA relied on structured input and output data from WHO VA 2016 questionnaires, and assigned CODs in WHO VA 2016 codes only. This required that these codes and forms be maintained with conversions between different form (e.g. WHO VA 2012 to WHO VA 2016) and code standards (e.g. WHO VA 2016 to ICD-10), which reduces interoperability and comparability with other incompatible models. GPT models did not require strict formats for training and testing data, which can capture latent and more ambiguous patterns (e.g. health-seeking behaviours and social issues) outside the scope of WHO VA codes and forms \citep{jebleeAutomaticallyDeterminingCause2019,kingQualityDiagnosticValue2016}. For example, GPT-3.5/4 had higher performance (+0.35-0.65 PCCC) than InterVA-5 and InSilicoVA for ambiguous ill-defined records across age groups. GPT models also performed better (+0.11-0.61 PCCC) on CODs with a rarer occurrence, such as nutritional deficiencies (n=11) and diabetes mellitus (n=27). Rarer CODs may be more difficult to capture by questionnaire due to lack of sample data, but it may possibly have richer contextual information from articles, web sources, or books that offer knowledge for GPT models to leverage.

@@

This section discusses research opportunities to improve GPT models for assigning CODs. Section \ref{discuss-opp-peng} discusses the potential to improve GPT models with prompt engineering and exploration of misclassified records, while Section \ref{discuss-opp-gsurveys} describes the application of GPT models for improving household surveys for better data quality. Section \ref{discuss-opp-cava} identifies an opportunity to integrate GPT, InterVA-5, and InSilicoVA models into VA systems for improving physician COD assignment.

@@

\subsubsection{Guided Household Surveys}\label{discuss-opp-gsurveys}

Finally, GPT models may also guide surveyors during VA interviews to probe households for better narrative information by suggesting better questions or questions that may have been missed by the surveyors. Lastly, as models can assign CODs on-demand, there is potential for models to provide immediate COD estimates during the data collection process to monitor data quality (e.g. comparing estimated to expected COD distributions for known areas as quality checks).

\subsubsection{Computer Assisted Verbal Autopsy}\label{discuss-opp-cava}
